---
title: "04E - dplyr 9: Row Operations and Binding"
author: "Piotr Kosowski"
date: "2026-03-09"
output: pdf_document
---

# Introduction

In earlier notebooks you learned filtering, transforming, grouping, and joining.
Another common task in real projects is managing rows directly:

1.  stacking data from multiple files,
2.  adding corrections,
3.  updating selected records,
4.  removing invalid records,
5.  checking if row-level changes are correct.

This notebook focuses on practical row operations in `dplyr` that are very useful in day-to-day analysis.

---

# Part I - Theory

## 1. When to use row operations instead of joins

Use row operations when your primary goal is to add, update, patch, or remove records.

A quick guide:

1.  use joins when you want to add columns from another table,
2.  use `bind_rows()` when you want to stack tables vertically,
3.  use `rows_*()` when you want database-like row changes by key.

---

## 2. Setup: small teaching tables

```{r}
library(dplyr)

jan_sales <- tibble(
  order_id = c(1001, 1002, 1003),
  customer_id = c("C01", "C02", "C03"),
  amount = c(120, 80, 150),
  status = c("paid", "pending", "paid")
)

feb_sales <- tibble(
  order_id = c(1004, 1005, 1006),
  customer_id = c("C01", "C04", "C05"),
  amount = c(95, 210, 60),
  status = c("paid", "paid", "returned")
)

jan_sales
feb_sales
```

These two tables have the same structure, so they are perfect for row binding.

---

## 3. `bind_rows()` - stack tables vertically

`bind_rows()` adds rows from one table under another.

```{r}
all_sales <- bind_rows(jan_sales, feb_sales)
all_sales
```

If columns are not identical, `bind_rows()` creates missing values (`NA`) where needed.

```{r}
feb_sales_extra <- feb_sales %>%
  mutate(channel = c("online", "store", "online"))

bind_rows(jan_sales, feb_sales_extra)
```

---

## 4. Track data source with `.id`

When combining many datasets, it is useful to keep source information.

```{r}
bind_rows(
  January = jan_sales,
  February = feb_sales,
  .id = "source_month"
)
```

This creates a new column that remembers where each row came from.

---

## 5. `bind_cols()` - combine by position (careful)

`bind_cols()` joins tables side by side by row position, not by key.

```{r}
order_flags <- tibble(
  is_large = c(FALSE, FALSE, TRUE),
  needs_review = c(FALSE, TRUE, FALSE)
)

bind_cols(jan_sales, order_flags)
```

Important warning:

1.  row counts must match,
2.  row order must already be aligned,
3.  if order is not aligned, result can be incorrect.

In many analytic tasks, key-based joins are safer than `bind_cols()`.

---

## 6. `rows_append()` - add new rows

`rows_append()` adds rows from `y` to `x`.

```{r}
customers_master <- tibble(
  customer_id = c("C01", "C02", "C03"),
  email = c("anna@example.com", "bartek@example.com", NA),
  city = c("Krakow", "Gdansk", "Warsaw")
)

new_customers <- tibble(
  customer_id = c("C04", "C05"),
  email = c("ewa@example.com", "filip@example.com"),
  city = c("Poznan", "Lodz")
)

rows_append(customers_master, new_customers)
```

Use this when new records should be added.

---

## 7. `rows_update()` - update existing rows by key

`rows_update()` modifies rows in `x` that match keys from `y`.

```{r}
email_updates <- tibble(
  customer_id = c("C02"),
  email = c("bartek.new@example.com"),
  city = c("Gdansk")
)

rows_update(customers_master, email_updates, by = "customer_id")
```

Good practice:

1.  keep keys unique in update table,
2.  update only fields you really want to replace.

---

## 8. `rows_patch()` - fill missing values only

`rows_patch()` updates only `NA` values in `x`.

```{r}
patch_data <- tibble(
  customer_id = c("C03"),
  email = c("celina@example.com"),
  city = c("Warsaw")
)

rows_patch(customers_master, patch_data, by = "customer_id")
```

This is useful when you want to complete missing values without overwriting existing non-missing data.

---

## 9. `rows_upsert()` - update if exists, insert if new

`rows_upsert()` is a practical "update or insert" operation.

```{r}
upsert_data <- tibble(
  customer_id = c("C02", "C06"),
  email = c("bartek.final@example.com", "gosia@example.com"),
  city = c("Gdansk", "Wroclaw")
)

rows_upsert(customers_master, upsert_data, by = "customer_id")
```

Result:

1.  `C02` is updated,
2.  `C06` is inserted.

---

## 10. `rows_delete()` - remove rows by key

`rows_delete()` removes rows whose keys appear in `y`.

```{r}
to_remove <- tibble(customer_id = c("C01"))

rows_delete(customers_master, to_remove, by = "customer_id")
```

Use carefully and keep deletion logic explicit.

---

## 11. Basic validation after row operations

Always check outcomes after row changes.

```{r}
updated_customers <- rows_upsert(customers_master, upsert_data, by = "customer_id")

updated_customers %>%
  summarise(
    rows_total = n(),
    unique_ids = n_distinct(customer_id),
    duplicated_ids = n() - n_distinct(customer_id)
  )
```

Also inspect unexpected keys with `anti_join()` when needed.

```{r}
upsert_data %>%
  anti_join(customers_master, by = "customer_id")
```

This shows rows that were new (not in original table).

---

## 12. Common beginner mistakes and checklist

Frequent mistakes:

1.  using `bind_cols()` when key-based logic is needed,
2.  forgetting key uniqueness before `rows_update()`/`rows_upsert()`,
3.  applying updates without checking row counts,
4.  deleting rows without keeping reproducible code.

Checklist before final output:

1.  Is the operation type correct (`bind`, `join`, or `rows_*`)?
2.  Is the key column clear and unique where required?
3.  Are row counts and key uniqueness checked after operation?
4.  Is every change traceable in code?

---

# Part II - Exercises

Create small tibbles in each task. Keep code explicit and readable.

## Exercise 1 - Stack two monthly tables with `bind_rows()`

**Goal:** combine same-structure tables.

**Task:** create `month1` and `month2` with columns `id`, `amount`; then stack them.

```{r}
# Your code here
# month1 with data
month1 <- data.frame(
  id = c(1, 2, 3),
  amount = c(100.50, 200.00, 150.75)
)

# month2 with data
month2 <- data.frame(
  id = c(1, 2, 4),
  amount = c(110.00, 200.00, 300.25)
)

bind_rows(month1, month2)
```

---

## Exercise 2 - `bind_rows()` with non-identical columns

**Goal:** see how missing columns become `NA`.

**Task:** add extra column `channel` only to second table and stack both tables.

```{r}
# Your code here
# month1 with data
month1 <- data.frame(
  id = c(1, 2, 3),
  amount = c(100.50, 200.00, 150.75)
)

# month2 with data
month2 <- data.frame(
  id = c(1, 2, 4),
  amount = c(110.00, 200.00, 300.25),
  channel = c(1, 2, 3)
)

bind_rows(month1, month2)
```

---

## Exercise 3 - Track source with `.id`

**Goal:** keep origin of each row.

**Task:** bind two named tables and create source column `source_batch`.

```{r}
# Your code here
# month1 with data
month1 <- data.frame(
  id = c(1, 2, 3),
  amount = c(100.50, 200.00, 150.75)
)

# month2 with data
month2 <- data.frame(
  id = c(1, 2, 4),
  amount = c(110.00, 200.00, 300.25)
)

bind_rows(
  Month1 = month1,
  Month2 = month2,
  .id = "source_month"
)
```

---

## Exercise 4 - Side-by-side combination with `bind_cols()`

**Goal:** combine aligned tables by position.

**Task:** create a 3-row sales table and a 3-row flags table, then bind columns.

```{r}
# Your code here
sales <- data.frame(
  id = c(101, 102, 103),
  product = c("Widget A", "Widget B", "Widget C"),
  amount = c(45.99, 120.00, 89.50),
  date = as.Date(c("2024-01-01", "2024-01-02", "2024-01-03"))
)

flags <- data.frame(
  id = c(101, 102, 103),
  is_active = c(TRUE, TRUE, FALSE),
  is_priority = c(FALSE, TRUE, FALSE),
  status_code = c("OK", "OK", "REVIEW")
)

bind_cols(sales, flags)
```

---

## Exercise 5 - Add new rows with `rows_append()`

**Goal:** append new records to master table.

**Task:** create customer master and new customers table; append rows.

```{r}
# Your code here
customer_master <- data.frame(
  cust_id = c(1001, 1002, 1003),
  name = c("Alice Johnson", "Bob Smith", "Charlie Davis"),
  email = c("alice@example.com", NA, "charlie@example.com"),
  total_spend = c(450.25, 1200.00, 0.00),
  stringsAsFactors = FALSE
)

new_customers <- data.frame(
  cust_id = c(1004, 1005, 1006),
  name = c("Diana Prince", "Edward Norton", "Fiona Gallagher"),
  email = c("diana@example.com", "edward@example.com", "fiona@example.com"),
  total_spend = c(50.00, 0.00, 85.20),
  stringsAsFactors = FALSE
)

rows_append(customer_master, new_customers)
customer_master
```

---

## Exercise 6 - Update existing rows with `rows_update()`

**Goal:** modify selected rows by key.

**Task:** update email for one existing customer using key `customer_id`.

```{r}
# Your code here
update_row <- tibble(
  cust_id = c(1001),
  email = c("alice_update@gmail.com")
)

rows_update(customer_master, update_row, by = "cust_id")
```

---

## Exercise 7 - Patch missing values with `rows_patch()`

**Goal:** fill missing fields without overwriting non-missing data.

**Task:** in master table with missing email, patch only that missing value.

```{r}
# Your code here
patch_row <- tibble(
  cust_id = c(1002),
  email = c("patched_email@gmail.com")
)

rows_patch(customer_master, patch_row, by = "cust_id")
```

---

## Exercise 8 - Update or insert with `rows_upsert()`

**Goal:** handle mixed update+new data.

**Task:** upsert one existing and one new customer by `customer_id`.

```{r}
# Your code here
customer_master

upsert_data <- tibble(
  cust_id = c(1001, 1004),
  email = c("updated_email@gmail.com", "new_email@gmail.com")
)

rows_upsert(customer_master, upsert_data, by = "cust_id")
```

---

## Exercise 9 - Remove selected rows with `rows_delete()`

**Goal:** remove rows by key list.

**Task:** delete one customer from table using key column.

```{r}
# Your code here
row_to_delete <- tibble(
  cust_id = c(1001)
)

rows_delete(customer_master, row_to_delete, by = "cust_id")
```

---

## Exercise 10 - Post-operation validation summary

**Goal:** verify integrity after row changes.

**Task:** after an upsert, compute row count, distinct key count, and number of duplicate keys.

```{r}
# Your code here
update_customers <- rows_upsert(customers_master, upsert_data, by = "cust_id")

updated_customers %>%
  summarise(
    rows_total = n(),
    unique_ids = n_distinct(cust_id),
    duplicated_ids = n() - n_distinct(cust_id)
  )
```

---

## Exercise 11 - Detect truly new keys

**Goal:** identify rows that were not in original table.

**Task:** use `anti_join()` to find keys present in update table but absent in master table.

```{r}
# Your code here
```

---

## Exercise 12 - Integration pipeline

**Goal:** combine row operations in one workflow.

**Task:** create sales data from two months and corrections table; then:

1.  stack monthly tables,
2.  apply `rows_update()` to correct one existing order,
3.  apply `rows_append()` to add one late order,
4.  verify final row count and unique order count.

```{r}
# Your code here
```

---

# Summary

In this final dplyr notebook you practiced practical row operations:

1.  stacking tables (`bind_rows`),
2.  source tracking (`.id`),
3.  careful side binding (`bind_cols`),
4.  database-like row changes (`rows_append`, `rows_update`, `rows_patch`, `rows_upsert`, `rows_delete`),
5.  validation checks after each operation.

These tools are essential when data changes over time and updates arrive in batches.

---

# References

1.  Wickham H, Francois R, Henry L, Muller K, Vaughan D. **dplyr reference**. https://dplyr.tidyverse.org/
2.  Wickham H, Cetinkaya-Rundel M, Grolemund G. **R for Data Science (2e)**, Data transformation chapters. https://r4ds.hadley.nz/
3.  Posit Cheatsheets. **Data transformation with dplyr**. https://posit.co/resources/cheatsheets/
4.  Wickham H. **Advanced R (2e)**, selected sections for deeper context. https://adv-r.hadley.nz/
