---
title: "Tornados Analysis (TidyTuesday) - Modeling"
author: "Leonardo Cárdenas Peralta, Daniel Enrique Molina Sanglade and Enrique Fernandez Salmeron"
date: "April-May 2026"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
    highlight: tango
---

# Introduction

In this section we build statistical models to go beyond description and answer concrete analytical questions about tornadoes in the United States.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(performance)
library(kableExtra)
library(broom)
library(dplyr)
library(ggplot2)
```

## Load the cleaned dataset

```{r load data, message=FALSE, warning=FALSE}
tornados_clean <- read_csv(here("Data", "tornados_clean.csv"))
glimpse(tornados_clean)

# Re-create the engineered columns from EDA (needed for modeling)
# New columns creation
tornados_clean <- tornados_clean %>%
  # 1. CALCULATED INDICATORS
  mutate(
    victims_total = inj + fat,                                  # 1. Total affected people
    is_deadly = fat > 0,                                        # 2. Binary death indictor
    area_sq_mi = (len * (wid / 1760)),                          # 3. Estimated area
    has_loss = !is.na(loss) & loss > 0                          # 4. Economic damage indicator
  ) %>%
  
  # 2. GROUPED CATEGORIES
  mutate(
    season = case_when(                                         # 5. Grouped by season
      mo %in% c(12, 1, 2) ~ 1,
      mo %in% c(3, 4, 5)  ~ 2,
      mo %in% c(6, 7, 8)  ~ 3,
      mo %in% c(9, 10, 11) ~ 4
    ),
    mag_category = case_when(                                   # 6. Agrupación de intensidad (Escala Fujita) Intensity grouping (Fujita scale)
      mag <= 1 ~ "Weak",
      mag <= 3 ~ "Strong",
      mag > 3  ~ "Violent",
      TRUE     ~ "Unknown"
    )
  ) %>%
  
  # 3. RECODED VARIABLES
  mutate(
    us_region = case_when(                                      # 7. State-region recoding
      st %in% c("CT", "ME", "MA", "NH", "RI", "VT", "NJ", "NY", "PA") ~ "Northeast",
      st %in% c("IL", "IN", "IA", "KS", "MI", "MN", "MO", "NE", "ND", "OH", "SD", "WI") ~ "Midwest",
      st %in% c("AL", "AR", "DE", "FL", "GA", "KY", "LA", "MD", "MS", "NC", "OK", "SC", "TN", "TX", "VA", "WV") ~ "South",
      TRUE ~ "West"
    )
  ) %>%
  
  # 4. TRANSFORMATIONS
  mutate(
    loss_log10 = ifelse(loss > 0, log10(loss), 0),              # 8. Logarithmic transformation of losses
    mag_factor = factor(mag, levels = 0:5, ordered = TRUE)      # 9. Mag factor ordered
  ) %>%
  
  # 5. AGGREGATED VALUES
  group_by(yr, st) %>%
  mutate(
    state_annual_fat_total = sum(fat, na.rm = TRUE),            # 10. Total death by state in a year
    state_annual_avg_mag = mean(as.numeric(mag), na.rm = TRUE)  # 11. Average mag by state in a year
  ) %>%
  ungroup() %>%
  
  # 6. EXTRA
  mutate(
    time_of_day = case_when(                                    # 12. Day-time grouping
      lubridate::hour(datetime_utc) >= 6 & lubridate::hour(datetime_utc) < 12 ~ "Morning",
      lubridate::hour(datetime_utc) >= 12 & lubridate::hour(datetime_utc) < 18 ~ "Afternoon",
      lubridate::hour(datetime_utc) >= 18 & lubridate::hour(datetime_utc) < 24 ~ "Evening",
      TRUE ~ "Night"
    )
  )

glimpse(tornados_clean)

```

---

# Research Questions
1- IS THERE A SIGNIFICANT RELATIONSHIP BETWEEN TWO VARIABLES?
We will observe if there is a relationship between the "magnitude" and the "victims_total" (injuries+fatalities)
```{r}
# We are going to take less rows, not the 68000 that it has.
# For example the first 5000
tornados_plot_5000 <- tornados_clean %>%
  slice_head(n = 5000)

tornados_plot_5000
```

```{r}
# We use "jitter" intead of "geom_point" to avoid overlaping data on the chart (jitter adds a little noise in each point to move them away from others) to allow the user to analyze it better
mag_victims <- tornados_plot_5000 %>% ggplot(aes(x = mag, y = victims_total)) +
  geom_jitter(alpha = 0.4, color = "steelblue")
mag_victims
```
```{r}
# Now, we will make antoher type of chart to can compare better at one sight.
# We will have all the victims of each type of "mag" and represent it with bars to see if it is ascending as we can imagine (as more "mag", more "victims", we could think)
victimas_por_magnitud <- tornados_plot_5000 %>%
  group_by(mag) %>%
  summarise(total_victimas = sum(victims_total, na.rm = TRUE)) %>%
  # Opcional: eliminar posibles magnitudes extrañas si las hay
  filter(!is.na(mag))

victimas_por_magnitud
```
```{r}
ggplot(victimas_por_magnitud, aes(x = factor(mag), y = total_victimas, fill = factor(mag))) +
  geom_col() +
  # Añadimos etiquetas de texto arriba de las barras para que se vea profesional
  geom_text(aes(label = total_victimas), vjust = -0.5) + 
  scale_fill_brewer(palette = "Reds") + # Un degradado de rojos (más peligro = más rojo)
  labs(
    title = "Total sum of victims by Tornado Magnitude",
    subtitle = "Data from 1950-2022",
    x = "Mag (Scale F/EF)",
    y = "Total sum of victims (Injuries + Fatalities)",
    fill = "Magnitude"
  ) +
  theme_minimal()
```

```{r}
# We can see there is some relation as the bars starts ascending, although the mag 4 is the biggest when we could think before knowing the chart, that it would be the 5th.

# This could be because, by chance one tornado of magnitude "4" was so deadly and distrubes the chart. Lets see if it happens


worse_tornados <- tornados_clean %>%
  group_by(mag) %>%
  summarise(max_victimas = max(victims_total, na.rm = TRUE)) %>%
  filter(!is.na(mag))

# Ver los récords en la consola
worse_tornados


# We show it in a chart
ggplot(worse_tornados, aes(x = factor(mag), y = max_victimas)) +
  # The line that goes from floor to the point
  geom_segment(aes(x = factor(mag), xend = factor(mag), y = 0, yend = max_victimas), 
               color = "grey") +
  # The point
  geom_point(aes(color = factor(mag)), size = 5) +
  # Exact number for no space to doubts
  geom_text(aes(label = max_victimas), vjust = -1.5, size = 4) +
  # Expand the Y axis to not cut
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Record casualties in a single tornado by magnitude",
  subtitle = "The 'worst-case scenario' recorded for each category",
  x = "Magnitude",
  y = "Maximum casualties in a single event",
  color = "Magnitude"
  ) +
  theme_minimal()
```
```{r}
# Now that we have manually selected 2 variables to see if they have some relation, we are going to search for those that have strong relation
library(corrr) # Paquete excelente para esto
library(ggcorrplot)
# We select the numeric cols that we are interested in
tornados_corr <- tornados_clean %>%
  select(mag, len, wid, loss, inj, fat, victims_total, yr, mo) %>%
  drop_na() # The correlation stops working if there are NA`s

# 2. Calculate the correlation matrix
corr_matrix <- cor(tornados_corr)

# This matrix, shows the relation between all the possible variables of the dataset (in this case we selected some of them, not all)
# It has values between -1 and 1.
# If its 1 ---> Means that both variables change the same if the other does
# If its -1 ---> Means that if one rises the other goes down
# If its 0 ---> Dont have any type of relation

# 3. Draw the heat map
ggcorrplot(corr_matrix, 
           hc.order = TRUE, 
           type = "lower",
           lab = TRUE, 
           colors = c("#6D9EC1", "white", "#E46726"),
           title = "Correlation Matrix: Tornado Variables",
           ggtheme = theme_minimal())

# We can see some relations following the color scale.
```

2- CAN ONE VARIABLE BE PREDICTED BASED ON OTHERS?
```{r}
library(tidyverse)
library(scales) # Required for better axis labels

# --- 1. DATA PREPARATION ---
# We create a specific subset for modeling to avoid issues with missing values (NAs)
# and focused on physical predictors (mag, len, wid) vs impact (loss)
data_model <- tornados_clean %>%
  select(loss, mag, len, wid) %>%
  drop_na()

# --- 2. LOGARITHMIC MODEL ---
# Why log? Because tornado damage is exponential. 
# A magnitude 5 is thousands of times more destructive than a magnitude 1.
# log(loss + 1) flattens these extremes so the linear model can "see" the trend.
model_log <- lm(log(loss + 1) ~ mag + len + wid, data = data_model)

# Print the statistical summary
# LOOK FOR: 'Adjusted R-squared' (model quality) and 'Pr(>|t|)' (significance)
summary(model_log)

# In this case, the error will be higher because we did not ignored the 0 loss tornados and disturbed the model
```

```{r}
# --- 3. FILTERED MODEL (Impact analysis) ---
# We filter out zero-loss cases to see how variables behave when damage actually occurs.
data_damage_only <- data_model %>% filter(loss > 0)
model_damage <- lm(log(loss) ~ mag + len, data = data_damage_only)

summary(model_damage)
```

```{r}
# --- 4. VISUALIZATION ---
# Using scale_y_log10 is crucial. Without it, the F5 tornadoes would 
# "crush" all other data points at the bottom of the graph.
ggplot(data_damage_only, aes(x = mag, y = loss)) +
  geom_jitter(color = "steelblue", alpha = 0.3, width = 0.2) + # Jitter reveals data density
  geom_smooth(method = "lm", color = "firebrick", se = TRUE) + # The red line is our prediction
  scale_y_log10(labels = scales::dollar_format()) + # Log scale for the Y axis
  labs(
    title = "Predictive Model: Tornado Magnitude vs. Economic Loss",
    subtitle = "Trend line calculated using Linear Regression (Log scale)",
    x = "Tornado Magnitude (F/EF Scale)",
    y = "Economic Loss (Logarithmic Scale)",
    caption = "Data: 1950-2022 | Filtered: losses > 0"
  ) +
  theme_minimal()
```

3- WHICH FACTORS SEEM TO MATTER MOST FOR A SELECTED OUTCOME?
```{r}
# In the previous summary of data, we have a clear example o fwhich atribute is more important in an outcome (in this case inn "losses")

# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 9.134522   0.016992   537.6   <2e-16 ***
# mag         1.358654   0.013264   102.4   <2e-16 ***
# len         0.041670   0.001255    33.2   <2e-16 ***

# The "t value" represents the "importance" of that atribute on the outcome.
# In this case, mag is the most important (following the results of the damage model (the second one))
```


```{r}
# SECOND PART

# Then build at least two different models using methods discussed during the course.
# Possible examples include:
# linear regression,
# logistic regression,
# or other relevant methods introduced in class.
```

```{r}
# We will make a LINEAR REGRESSION on predict the amount of LOSS
library(tidymodels)
library(tidyverse)

tornados_modeling <- tornados_clean %>%
  dplyr::select(loss_log10, mag, len, wid, season, area_sq_mi) %>%
  drop_na()

# ==========================================
# 2. DATA SPLITTING (rsample)
# ==========================================
tornado_split <- initial_split(tornados_modeling, prop = 0.8)
tornado_train <- training(tornado_split)
tornado_test  <- testing(tornado_split)

# ==========================================
# 3. RECIPE SPECIFICATION (recipes)
# ==========================================
# Like in the notebooks, define the formula and normalize
lm_recipe <- recipe(loss_log10 ~ mag + len + wid + season + area_sq_mi, data = tornado_train) %>%
  step_normalize(all_predictors())

# ==========================================
# 4. MODEL SPECIFICATION (parsnip)
# ==========================================
lm_spec <- linear_reg() %>%
  set_engine("lm")

# ==========================================
# 5. WORKFLOW CREATION (workflows)
# ==========================================
lm_wflow <- workflow() %>%
  add_recipe(lm_recipe) %>%
  add_model(lm_spec)

# ==========================================
# 6. MODEL FITTING
# ==========================================
lm_fit <- fit(lm_wflow, data = tornado_train)

# ==========================================
# 7. EVALUATION (yardstick)
# ==========================================
# We make predicitins in the test data as in the exercises
lm_predictions <- predict(lm_fit, new_data = tornado_test) %>%
  bind_cols(tornado_test)

# Calculate the regresion data (RMSE y R-squared)
metrics_lm <- metrics(lm_predictions, truth = loss_log10, estimate = .pred)
print(metrics_lm)
```
```{r}
# Explication of this data:
#"rmse" and "mae" are error metrics
# "rmse" first error^2, then square and then average
# "mae" its simiple. Sums all the error commited and makes an average
# "rsq" its 28%, and means that our atriburtes explain the 28% of variability of the results
```

```{r}
library(ggplot2)

# Chart of the predictions of the test
ggplot(lm_predictions, aes(x = loss_log10, y = .pred)) +
  geom_point(alpha = 0.2, color = "darkblue") +
  # Add the ideal referencial line (Precition = Reality)
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(
   title = "Linear Model Evaluation: Predictions vs Reality",
    subtitle = "The red dashed line represents the perfect prediction",
    x = "Actual Value (loss_log10)",
    y = "Model Prediction (.pred)"
  ) +
  theme_minimal()
```
```{r}
# When more blue points (reality cases of test) are closer to the red line (prediction), better is the model
```

```{r}
lm_fit %>% 
  extract_fit_parsnip() %>% 
  tidy()
```
```{r}
# `estimate`**: The weight (coefficient). It indicates how much the model's prediction changes for each one-unit increase in that specific variable.

# `std.error`**: The standard error of the estimate. It measures the margin of error for that weight; the smaller it is, the more precise the model's calculation.

# `statistic`**: The t-statistic (or t-value), which serves as our "importance ranking." It is calculated by dividing the `estimate` by the `std.error`.

# `p.value`**: The statistical significance threshold. If this value is below 0.05 (in your output it displays as 0, meaning it is infinitely small), the variable is 100% statistically significant and the relationship is not due to random chance.
```

```{r}
library(vip)

# Extract the inner model and shows the importance of the used atributes to predict the outcome of losses
lm_fit %>%
  extract_fit_parsnip() %>%
  vip(geom = "col", aesthetics = list(fill = "darkorange")) +
  labs(title = "Importancia de las Variables en el Daño Económico") +
  theme_minimal()
```

```{r}
# We will make a LOGISTIC REGRESSION on predict if the tornado will cause any VICTIM
# ==========================================
# 1. SETTING UP DATA FOR CLASSIFICATION
# ==========================================
tornados_class <- tornados_clean %>%
  # Convertimos el campo lógico en factor ("TRUE" -> tipo categoría)
  mutate(is_deadly = as.factor(is_deadly)) %>% 
  dplyr::select(is_deadly, mag, len, wid, season, area_sq_mi) %>%
  drop_na()

# ==========================================
# 2. DATA SPLITTING
# ==========================================
set.seed(123)
class_split <- initial_split(tornados_class, prop = 0.8)
class_train <- training(class_split)
class_test  <- testing(class_split)

# ==========================================
# 3. RECIPE SPECIFICATION
# ==========================================
# Definimos que queremos predecir is_deadly
logit_recipe <- recipe(is_deadly ~ mag + len + wid + season + area_sq_mi, data = class_train) %>%
  step_normalize(all_predictors())

# ==========================================
# 4. MODEL SPECIFICATION (Aquí cambia a clasificación)
# ==========================================
logit_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification") # ¡Paso clave de tus apuntes!

# ==========================================
# 5. WORKFLOW CREATION
# ==========================================
logit_wflow <- workflow() %>%
  add_recipe(logit_recipe) %>%
  add_model(logit_spec)

# ==========================================
# 6. MODEL FITTING
# ==========================================
logit_fit <- fit(logit_wflow, data = class_train)

# ==========================================
# 7. EVALUATION (Matriz de confusión y Accuracy)
# ==========================================
# En clasificación tus notebooks piden predecir la clase (.pred_class)
logit_predictions <- predict(logit_fit, new_data = class_test) %>%
  bind_cols(class_test)

# Evaluamos con la matriz de confusión exacta que te piden en los ejercicios
conf_matrix <- logit_predictions %>%
  conf_mat(truth = is_deadly, estimate = .pred_class)

print(conf_matrix)

# Calculamos la precisión global (Accuracy)
accuracy_logit <- logit_predictions %>%
  accuracy(truth = is_deadly, estimate = .pred_class)

print(accuracy_logit)
```
```{r}
# 1. Table with the COnufsion Matrix (hits and fails)
print(conf_matrix)

# 2. Chart more visible
conf_matrix %>% 
  autoplot(type = "heatmap") +
  labs(title = "Confusion Matrix - Logistic Model (is_deadly)")
```


1. **Can we predict the total number of victims of a tornado** (injuries + fatalities) based on its characteristics (magnitude, path length, width, season, area)?  
   → Answered with **Linear Regression**.

2. **Can we predict the total cost of a tornado** based on its characteristics (magnitude, path length, width, season, area)?
   → Answered with **Linear Regression**.

3. **Can we predict whether a tornado will be deadly** (at least one fatality) based on its characteristics?  
   → Answered with **Logistic Regression**.

---

# Data Preparation for Modeling

```{r model data, message=FALSE, warning=FALSE}
# Select and filter the variables we will use
model_data <- tornados_clean %>%
  dplyr::select(
    victims_total,
    is_deadly,
    mag,
    len,
    wid,
    area_sq_mi,
    season,
    us_region,
    mag_category
  ) %>%
  filter(
    !is.na(victims_total),
    !is.na(mag),
    !is.na(len),
    !is.na(wid),
    !is.na(season),
    !is.na(us_region)
  )

cat("Rows available for modeling:", nrow(model_data), "\n")
cat("Columns used:", ncol(model_data), "\n")
```

---

# First models — Linear Regression

## Victims Analysis

First we are going to see how the victmis are distributed along the different ranges

```{r}
ggplot(tornados_clean, aes(x = victims_total)) + 
  geom_histogram() + 
  scale_y_log10()
```

## Why these variables?

We want to predict `victims_total` (injuries + fatalities), which is a **continuous numeric variable**. We use:

- `mag` → tornado magnitude on the Fujita scale: more powerful tornadoes are expected to cause more damage.
- `len` → path length in miles: longer paths affect more area and more people.
- `wid` → width in yards: wider tornadoes cover more surface.
- `season` → time of year: tornadoes in Spring tend to be more intense.
- `us_region` → geographic region: the South and Midwest are more densely affected.

## How the model works

Linear regression finds the equation:

```
victims_total = β0 + β1*mag + β2*len + β3*wid + β4*season + β5*us_region + ε
```

Each coefficient (β) tells us how much `victims_total` changes for each unit increase in that variable, holding the rest constant.

```{r linear model, message=FALSE, warning=FALSE}
library(MASS) # Required for glm.nb

# Model 1: Linear Regression
model_linear <- lm(
  victims_total ~ mag + len + wid + season + us_region,
  data = model_data
)

# Model 2: Poisson Regression
model_poisson <- glm(
  victims_total ~ mag + len + wid + season + us_region,
  family = poisson(link = "log"),
  data = model_data
)

# Model 3: Negative Binomial Regression
model_negbinom <- glm.nb(
  victims_total ~ mag + len + wid + season + us_region,
  data = model_data
)

# Model comparison
performance_comparison <- performance::compare_performance(
  model_linear, model_poisson, model_negbinom, 
  rank = TRUE
)

summary(model_linear)
summary(model_poisson)
summary(model_negbinom)
```

## Coefficient interpretation

```{r linear coef, message=FALSE, warning=FALSE}
tidy(model_linear) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes**", "No")
  ) %>%
  kable(digits = 4, caption = "Linear Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

tidy(model_poisson) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes**", "No")
  ) %>%
  kable(digits = 4, caption = "Poisson Regression Coefficients (Log-Link)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

tidy(model_negbinom) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes**", "No")
  ) %>%
  kable(digits = 4, caption = "Negative Binomial Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

**How to read the table:**

- `estimate` → for each unit increase in that variable, victims_total changes by this amount.
- `std.error` → the variance of the estiamation. The smaller it is with respect to the estimate the better.
- `statistic` → how much bigger is the estimate than the error. Above 2 is a good indicator.
- `p.value` → if below 0.05, the variable has a statistically significant effect.

- Variables marked **Yes** are meaningful predictors and in **No** variables there is not enough statistical evidence.
- For reference: in the group of season terms, the comparison is done with respect to Fall.
- For reference: in the group of region terms, the comparison is done with respect to Midwest.

## Model quality

```{r linear quality, message=FALSE, warning=FALSE}
glance(model_linear) %>%
  dplyr::select(AIC, BIC, deviance, logLik, df.residual) %>%
  kable(digits = 4, caption = "Linear Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

glance(model_poisson) %>%
  dplyr::select(AIC, BIC, deviance, logLik, df.residual) %>%
  kable(digits = 4, caption = "Poisson Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

glance(model_negbinom) %>%
  dplyr::select(AIC, BIC, deviance, logLik, df.residual) %>%
  kable(digits = 4, caption = "Negative Binomial Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

```{r}
compare_performance(model_linear, model_poisson, model_negbinom) %>%
  kable(digits = 4, caption = "Global Model Comparison (Linear vs Poisson vs NegBin)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**How to read:**

- `AIC (Akaike Information Criterion)` estimates the relative amount of infomration lost by a model (the lower the better).
- `AIC_wt` the best model to pick.
- `BIC (Bayesian Information Cirterio)` simialr to AIC but with stronger penalty for including additional parameters (the lower the better).
- `BIC_wt` the best model to pick.
- `RMSE (Root Mean Square Error)` gives error at the prediction with respect to the real values observed.
- `Sigma` standard deviation
- `R2_Nagelkerke` is and adaptation of R2 of linear models to GLM (Generalised Linear Model). The lower the better.
- `R2 / R2_adjusted` explains how good is the model and its "goodness-of-fit".

## Diagnostic plots

```{r linear diagnostics, message=FALSE, warning=FALSE}
diag_plot <- performance::check_model(model_linear)

diag_plot_visual <- plot(diag_plot)

ggsave(
  filename = here("Outputs", "lin_reg_model_diagnostics.png"),
  plot = diag_plot_visual,
  width = 12,
  height = 10,
  dpi = 1000
)

diag_poisson <- performance::check_model(model_poisson)
p_diag_poisson <- plot(diag_poisson)

ggsave(
  filename = here("Outputs", "poisson_diagnostics.png"),
  plot = p_diag_poisson,
  width = 15, height = 12, dpi = 1000
)

diag_negbinom <- performance::check_model(model_negbinom)
p_diag_negbinom <- plot(diag_negbinom)

ggsave(
  filename = here("Outputs", "negbinom_diagnostics.png"),
  plot = p_diag_negbinom,
  width = 15, height = 12, dpi = 1000
)

```

# Model 2 — Logistic Regression

## Why these variables?

We want to predict `is_deadly` (TRUE/FALSE — did the tornado kill anyone?), which is a **binary variable**. This requires logistic regression instead of linear regression.

We use the same predictors as before since they describe the tornado's physical intensity, which is directly related to its lethality.

## How the model works

Logistic regression models the **probability** that `is_deadly = TRUE`:

```
log(P/(1-P)) = β0 + β1*mag + β2*len + β3*wid + β4*season + β5*us_region
```

The output is always a probability between 0 and 1. We use `family = binomial` to tell R this is logistic regression.

```{r logistic model, message=FALSE, warning=FALSE}
model_logistic <- glm(
  is_deadly ~ mag + len + wid + season + us_region,
  data = model_data,
  family = binomial
)

summary(model_logistic)
```

## Coefficient interpretation (Odds Ratios)

In logistic regression, coefficients are interpreted as **odds ratios** after applying `exp()`:

```{r logistic coef, message=FALSE, warning=FALSE}
tidy(model_logistic) %>%
  mutate(
    odds_ratio  = exp(estimate),
    significant = ifelse(p.value < 0.05, "Yes ***", "No")
  ) %>%
  dplyr::select(term, estimate, odds_ratio, p.value, significant) %>%
  kable(digits = 4, caption = "Logistic Regression — Odds Ratios") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**How to read odds ratios:**

- `odds_ratio > 1` → that variable **increases** the probability of a deadly tornado.
- `odds_ratio < 1` → that variable **decreases** the probability.
- `odds_ratio = 1` → no effect.

For example, if `mag` has an odds ratio of 2.5, each additional Fujita scale point multiplies the odds of a deadly tornado by 2.5.

## Model quality

```{r logistic quality, message=FALSE, warning=FALSE}
glance(model_logistic) %>%
  dplyr::select(null.deviance, deviance, AIC, BIC) %>%
  kable(digits = 2, caption = "Logistic Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**How to read:**

- `AIC` / `BIC` → lower is better. Used to compare models of the same type.
- `deviance` → how well the model fits. Lower deviance vs `null.deviance` means improvement over a model that predicts nothing.

```{r logistic accuracy, message=FALSE, warning=FALSE}
# Predicted probabilities and classification
model_data <- model_data %>%
  mutate(
    pred_prob    = predict(model_logistic, type = "response"),
    pred_deadly  = pred_prob > 0.5
  )

# Confusion matrix
conf_matrix <- table(
  Predicted = model_data$pred_deadly,
  Actual    = model_data$is_deadly
)

conf_matrix %>%
  kable(caption = "Confusion Matrix — Logistic Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Overall accuracy
accuracy <- mean(model_data$pred_deadly == model_data$is_deadly, na.rm = TRUE)
cat("Model accuracy:", round(accuracy * 100, 2), "%\n")
```

**How to read the confusion matrix:**

- **True Positives** (top-left): correctly predicted deadly tornadoes.
- **True Negatives** (bottom-right): correctly predicted non-deadly tornadoes.
- **False Positives**: predicted deadly but wasn't.
- **False Negatives**: predicted not deadly but it was.

---

# Model Comparison

```{r comparison, message=FALSE, warning=FALSE}
comparison_full <- data.frame(
  Model              = c("Linear Regression", "Logistic Regression", "Poisson Regression", "Negative Binomial"),
  Dependent_Variable  = c("log(victims_total + 1)", "is_deadly (binary)", "victims_total (counts)", "victims_total (counts)"),
  R_Function         = c("lm()", "glm(..., family=binomial)", "glm(..., family=poisson)", "glm.nb()"),
  Question_Answered  = c("How much does Y change?", "Will it be deadly (Yes/No)?", "What is the expected count?", "What is the count (with outliers)?"),
  Quality_Metric     = c("R² / Adj R²", "Accuracy / AIC", "AIC / Deviance", "AIC / BIC / Dispersion"),
  Interpretation     = c("Direct coefficients", "Odds Ratios", "Incident Rate Ratios", "Incident Rate Ratios")
)

comparison_full %>%
  kable(caption = "Comprehensive Comparison of Tornado Impact Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Conclusions

- **Linear Regression (Log-Transformed)**: This model serves as our baseline. By applying a log(y+1) transformation, we reveal how physical characteristics (magnitude, length, width) predict the scale of the impact. It is highly interpretable for general planning, telling us, for example, that a 1-unit increase in magnitude corresponds to a specific percentage increase in victims.
- **Poisson Regression**: This is our first specialized "count model." It assumes that the number of victims follows a Poisson distribution where the mean equals the variance. While it provides Incident Rate Ratios (IRR)—telling us how much each factor multiplies the expected number of victims—it often struggles with "Overdispersion" in tornado data (where rare, catastrophic events create more variance than the model expects).
- **Negative Binomial Regression**: This is the "Gold Standard" for our dataset. It functions like the Poisson model but includes an extra parameter to handle the extreme variance and "long tail" of deadly tornadoes. By successfully modeling the high number of zeros (non-deadly tornadoes) alongside rare mass-casualty events, it provides the most mathematically honest and accurate estimates of casualty counts.```
- **Logistic Regression**: Instead of predicting "how many," this model answers the binary question: "Will this tornado be deadly?" By calculating Odds Ratios ($exp(\beta)$), it allows us to classify risk. This is the most operationally useful tool for emergency services and early warning systems, where a "Yes/No" threshold is needed to trigger specific evacuation protocols.
