---
title: "Tornados Analysis (TidyTuesday) - Modeling"
author: "Leonardo Cárdenas Peralta, Daniel Enrique Molina Sanglade and Enrique Fernandez Salmeron"
date: "April-May 2026"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
    highlight: tango
---

# Introduction

In this section we build statistical models to go beyond description and answer concrete analytical questions about tornadoes in the United States.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(performance)
library(kableExtra)
library(broom)
```

## Load the cleaned dataset

```{r load data, message=FALSE, warning=FALSE}
tornados_clean <- read_csv(here("Data", "tornados_clean.csv"))

# Re-create the engineered columns from EDA (needed for modeling)
tornados_clean <- tornados_clean %>%
  mutate(
    victims_total  = inj + fat,
    is_deadly      = fat > 0,
    area_sq_mi     = len * (wid / 1760),
    has_loss       = !is.na(loss) & loss > 0,
    season = case_when(
      mo %in% c(12, 1, 2)  ~ "Winter",
      mo %in% c(3, 4, 5)   ~ "Spring",
      mo %in% c(6, 7, 8)   ~ "Summer",
      mo %in% c(9, 10, 11) ~ "Fall"
    ),
    mag_category = case_when(
      mag <= 1 ~ "Weak",
      mag <= 3 ~ "Strong",
      mag > 3  ~ "Violent",
      TRUE     ~ "Unknown"
    ),
    us_region = case_when(
      st %in% c("CT","ME","MA","NH","RI","VT","NJ","NY","PA") ~ "Northeast",
      st %in% c("IL","IN","IA","KS","MI","MN","MO","NE","ND","OH","SD","WI") ~ "Midwest",
      st %in% c("AL","AR","DE","FL","GA","KY","LA","MD","MS","NC","OK","SC","TN","TX","VA","WV") ~ "South",
      TRUE ~ "West"
    ),
    loss_log10 = ifelse(loss > 0, log10(loss), 0)
  )
```

---

# Research Questions

Based on what we observed in the EDA, we formulate two analytical questions:

1. **Can we predict the total number of victims of a tornado** (injuries + fatalities) based on its physical characteristics (magnitude, path length, width)?  
   → Answered with **Linear Regression**.

2. **Can we predict whether a tornado will be deadly** (at least one fatality) based on its characteristics?  
   → Answered with **Logistic Regression**.

---

# Data Preparation for Modeling

```{r model data, message=FALSE, warning=FALSE}
# Select and filter the variables we will use
model_data <- tornados_clean %>%
  dplyr::select(
    victims_total,
    is_deadly,
    mag,
    len,
    wid,
    area_sq_mi,
    season,
    us_region,
    mag_category
  ) %>%
  filter(
    !is.na(victims_total),
    !is.na(mag),
    !is.na(len),
    !is.na(wid),
    !is.na(season),
    !is.na(us_region)
  )

cat("Rows available for modeling:", nrow(model_data), "\n")
cat("Columns used:", ncol(model_data), "\n")
```

---

# First models — Linear Regression

## Victims Analysis

First we are going to see how the victmis are distributed along the different ranges

```{r}
ggplot(tornados_clean, aes(x = victims_total)) + 
  geom_histogram() + 
  scale_y_log10()
```

## Why these variables?

We want to predict `victims_total` (injuries + fatalities), which is a **continuous numeric variable**. We use:

- `mag` → tornado magnitude on the Fujita scale: more powerful tornadoes are expected to cause more damage.
- `len` → path length in miles: longer paths affect more area and more people.
- `wid` → width in yards: wider tornadoes cover more surface.
- `season` → time of year: tornadoes in Spring tend to be more intense.
- `us_region` → geographic region: the South and Midwest are more densely affected.

## How the model works

Linear regression finds the equation:

```
victims_total = β0 + β1*mag + β2*len + β3*wid + β4*season + β5*us_region + ε
```

Each coefficient (β) tells us how much `victims_total` changes for each unit increase in that variable, holding the rest constant.

```{r linear model, message=FALSE, warning=FALSE}
library(MASS) # Required for glm.nb

# Model 1: Linear Regression
model_linear <- lm(
  victims_total ~ mag + len + wid + season + us_region,
  data = model_data
)

# Model 2: Poisson Regression
model_poisson <- glm(
  victims_total ~ mag + len + wid + season + us_region,
  family = poisson(link = "log"),
  data = model_data
)

# Model 3: Negative Binomial Regression
model_negbinom <- glm.nb(
  victims_total ~ mag + len + wid + season + us_region,
  data = model_data
)

# Model comparison
performance_comparison <- performance::compare_performance(
  model_linear, model_poisson, model_negbinom, 
  rank = TRUE
)

summary(model_linear)
summary(model_poisson)
summary(model_negbinom)
```

## Coefficient interpretation

```{r linear coef, message=FALSE, warning=FALSE}
tidy(model_linear) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes**", "No")
  ) %>%
  kable(digits = 4, caption = "Linear Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

tidy(model_poisson) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes**", "No")
  ) %>%
  kable(digits = 4, caption = "Poisson Regression Coefficients (Log-Link)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

tidy(model_negbinom) %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Yes**", "No")
  ) %>%
  kable(digits = 4, caption = "Negative Binomial Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

**How to read the table:**

- `estimate` → for each unit increase in that variable, victims_total changes by this amount.
- `std.error` → the variance of the estiamation. The smaller it is with respect to the estimate the better.
- `statistic` → how much bigger is the estimate than the error. Above 2 is a good indicator.
- `p.value` → if below 0.05, the variable has a statistically significant effect.

- Variables marked **Yes** are meaningful predictors and in **No** variables there is not enough statistical evidence.
- For reference: in the group of season terms, the comparison is done with respect to Fall.
- For reference: in the group of region terms, the comparison is done with respect to Midwest.

## Model quality

```{r linear quality, message=FALSE, warning=FALSE}
glance(model_linear) %>%
  dplyr::select(AIC, BIC, deviance, logLik, df.residual) %>%
  kable(digits = 4, caption = "Linear Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

glance(model_poisson) %>%
  dplyr::select(AIC, BIC, deviance, logLik, df.residual) %>%
  kable(digits = 4, caption = "Poisson Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

glance(model_negbinom) %>%
  dplyr::select(AIC, BIC, deviance, logLik, df.residual) %>%
  kable(digits = 4, caption = "Negative Binomial Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

```{r}
compare_performance(model_linear, model_poisson, model_negbinom) %>%
  kable(digits = 4, caption = "Global Model Comparison (Linear vs Poisson vs NegBin)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**How to read:**

- `AIC (Akaike Information Criterion)` estimates the relative amount of infomration lost by a model (the lower the better).
- `AIC_wt` the best model to pick.
- `BIC (Bayesian Information Cirterio)` simialr to AIC but with stronger penalty for including additional parameters (the lower the better).
- `BIC_wt` the best model to pick.
- `RMSE (Root Mean Square Error)` gives error at the prediction with respect to the real values observed.
- `Sigma` standard deviation
- `R2_Nagelkerke` is and adaptation of R2 of linear models to GLM (Generalised Linear Model). The lower the better.
- `R2 / R2_adjusted` explains how good is the model and its "goodness-of-fit".

## Diagnostic plots

```{r linear diagnostics, message=FALSE, warning=FALSE}
diag_plot <- performance::check_model(model_linear)

diag_plot_visual <- plot(diag_plot)

ggsave(
  filename = here("Outputs", "lin_reg_model_diagnostics.png"),
  plot = diag_plot_visual,
  width = 12,
  height = 10,
  dpi = 1000
)

diag_poisson <- performance::check_model(model_poisson)
p_diag_poisson <- plot(diag_poisson)

ggsave(
  filename = here("Outputs", "poisson_diagnostics.png"),
  plot = p_diag_poisson,
  width = 15, height = 12, dpi = 1000
)

diag_negbinom <- performance::check_model(model_negbinom)
p_diag_negbinom <- plot(diag_negbinom)

ggsave(
  filename = here("Outputs", "negbinom_diagnostics.png"),
  plot = p_diag_negbinom,
  width = 15, height = 12, dpi = 1000
)

```

# Model 2 — Logistic Regression

## Why these variables?

We want to predict `is_deadly` (TRUE/FALSE — did the tornado kill anyone?), which is a **binary variable**. This requires logistic regression instead of linear regression.

We use the same predictors as before since they describe the tornado's physical intensity, which is directly related to its lethality.

## How the model works

Logistic regression models the **probability** that `is_deadly = TRUE`:

```
log(P/(1-P)) = β0 + β1*mag + β2*len + β3*wid + β4*season + β5*us_region
```

The output is always a probability between 0 and 1. We use `family = binomial` to tell R this is logistic regression.

```{r logistic model, message=FALSE, warning=FALSE}
model_logistic <- glm(
  is_deadly ~ mag + len + wid + season + us_region,
  data = model_data,
  family = binomial
)

summary(model_logistic)
```

## Coefficient interpretation (Odds Ratios)

In logistic regression, coefficients are interpreted as **odds ratios** after applying `exp()`:

```{r logistic coef, message=FALSE, warning=FALSE}
tidy(model_logistic) %>%
  mutate(
    odds_ratio  = exp(estimate),
    significant = ifelse(p.value < 0.05, "Yes ***", "No")
  ) %>%
  dplyr::select(term, estimate, odds_ratio, p.value, significant) %>%
  kable(digits = 4, caption = "Logistic Regression — Odds Ratios") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**How to read odds ratios:**

- `odds_ratio > 1` → that variable **increases** the probability of a deadly tornado.
- `odds_ratio < 1` → that variable **decreases** the probability.
- `odds_ratio = 1` → no effect.

For example, if `mag` has an odds ratio of 2.5, each additional Fujita scale point multiplies the odds of a deadly tornado by 2.5.

## Model quality

```{r logistic quality, message=FALSE, warning=FALSE}
glance(model_logistic) %>%
  dplyr::select(null.deviance, deviance, AIC, BIC) %>%
  kable(digits = 2, caption = "Logistic Model Quality Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**How to read:**

- `AIC` / `BIC` → lower is better. Used to compare models of the same type.
- `deviance` → how well the model fits. Lower deviance vs `null.deviance` means improvement over a model that predicts nothing.

```{r logistic accuracy, message=FALSE, warning=FALSE}
# Predicted probabilities and classification
model_data <- model_data %>%
  mutate(
    pred_prob    = predict(model_logistic, type = "response"),
    pred_deadly  = pred_prob > 0.5
  )

# Confusion matrix
conf_matrix <- table(
  Predicted = model_data$pred_deadly,
  Actual    = model_data$is_deadly
)

conf_matrix %>%
  kable(caption = "Confusion Matrix — Logistic Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Overall accuracy
accuracy <- mean(model_data$pred_deadly == model_data$is_deadly, na.rm = TRUE)
cat("Model accuracy:", round(accuracy * 100, 2), "%\n")
```

**How to read the confusion matrix:**

- **True Positives** (top-left): correctly predicted deadly tornadoes.
- **True Negatives** (bottom-right): correctly predicted non-deadly tornadoes.
- **False Positives**: predicted deadly but wasn't.
- **False Negatives**: predicted not deadly but it was.

---

# Model Comparison

```{r comparison, message=FALSE, warning=FALSE}
comparison_full <- data.frame(
  Model              = c("Linear Regression", "Logistic Regression", "Poisson Regression", "Negative Binomial"),
  Dependent_Variable  = c("log(victims_total + 1)", "is_deadly (binary)", "victims_total (counts)", "victims_total (counts)"),
  R_Function         = c("lm()", "glm(..., family=binomial)", "glm(..., family=poisson)", "glm.nb()"),
  Question_Answered  = c("How much does Y change?", "Will it be deadly (Yes/No)?", "What is the expected count?", "What is the count (with outliers)?"),
  Quality_Metric     = c("R² / Adj R²", "Accuracy / AIC", "AIC / Deviance", "AIC / BIC / Dispersion"),
  Interpretation     = c("Direct coefficients", "Odds Ratios", "Incident Rate Ratios", "Incident Rate Ratios")
)

comparison_full %>%
  kable(caption = "Comprehensive Comparison of Tornado Impact Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Conclusions

- **Linear Regression (Log-Transformed)**: This model serves as our baseline. By applying a log(y+1) transformation, we reveal how physical characteristics (magnitude, length, width) predict the scale of the impact. It is highly interpretable for general planning, telling us, for example, that a 1-unit increase in magnitude corresponds to a specific percentage increase in victims.
- **Poisson Regression**: This is our first specialized "count model." It assumes that the number of victims follows a Poisson distribution where the mean equals the variance. While it provides Incident Rate Ratios (IRR)—telling us how much each factor multiplies the expected number of victims—it often struggles with "Overdispersion" in tornado data (where rare, catastrophic events create more variance than the model expects).
- **Negative Binomial Regression**: This is the "Gold Standard" for our dataset. It functions like the Poisson model but includes an extra parameter to handle the extreme variance and "long tail" of deadly tornadoes. By successfully modeling the high number of zeros (non-deadly tornadoes) alongside rare mass-casualty events, it provides the most mathematically honest and accurate estimates of casualty counts.```
- **Logistic Regression**: Instead of predicting "how many," this model answers the binary question: "Will this tornado be deadly?" By calculating Odds Ratios ($exp(\beta)$), it allows us to classify risk. This is the most operationally useful tool for emergency services and early warning systems, where a "Yes/No" threshold is needed to trigger specific evacuation protocols.
