---
title: "Tornados Analysis (TidyTuesday) - Modeling"
author: "Leonardo Cárdenas Peralta, Daniel Enrique Molina Sanglade and Enrique Fernandez Salmeron"
date: "April-May 2026"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
    highlight: tango
---

# Introduction

In this section we will do:
- Preparation of data (specially done in "eda")
- Search for relations between variables (we will notice a "problem" on the correlation-matrix method: PEARSON or SPEARMAN)
- Regression models and result analysis of data
- Comparassion and interpretation of model results and errors
- Multiple variable regression model and comparation with single ones
- Logistic regression (noticing some errors and correcting then, as balancing)

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(performance)
library(kableExtra)
library(broom)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggeffects)
library(cvms)
library(tibble)
```

## Load the cleaned dataset

```{r load data linear, message=FALSE, warning=FALSE}
tornados_clean <- read_csv(here("Data", "tornados_clean.csv"))

# Re-create the engineered columns from EDA (needed for modeling)
# New columns creation
tornados_clean <- tornados_clean %>%
  # 1. CALCULATED INDICATORS
  mutate(
    victims_total = inj + fat,                                  # 1. Total affected people
    is_deadly = fat > 0,                                        # 2. Binary death indictor
    area_sq_mi = (len * (wid / 1760)),                          # 3. Estimated area
    has_loss = !is.na(loss) & loss > 0                          # 4. Economic damage indicator
  ) %>%
  
  # 2. GROUPED CATEGORIES
  mutate(
    season = case_when(                                         # 5. Grouped by season
      mo %in% c(12, 1, 2) ~ 1,
      mo %in% c(3, 4, 5)  ~ 2,
      mo %in% c(6, 7, 8)  ~ 3,
      mo %in% c(9, 10, 11) ~ 4
    ),
    mag_category = case_when(                                   # 6. Agrupación de intensidad (Escala Fujita) Intensity grouping (Fujita scale)
      mag <= 1 ~ "Weak",
      mag <= 3 ~ "Strong",
      mag > 3  ~ "Violent",
      TRUE     ~ "Unknown"
    )
  ) %>%
  
  # 3. RECODED VARIABLES
  mutate(
    us_region = case_when(                                      # 7. State-region recoding
      st %in% c("CT", "ME", "MA", "NH", "RI", "VT", "NJ", "NY", "PA") ~ "Northeast",
      st %in% c("IL", "IN", "IA", "KS", "MI", "MN", "MO", "NE", "ND", "OH", "SD", "WI") ~ "Midwest",
      st %in% c("AL", "AR", "DE", "FL", "GA", "KY", "LA", "MD", "MS", "NC", "OK", "SC", "TN", "TX", "VA", "WV") ~ "South",
      TRUE ~ "West"
    )
  ) %>%
  
  # 4. TRANSFORMATIONS
  mutate(
    loss_log10 = ifelse(loss > 0, log10(loss), 0),              # 8. Logarithmic transformation of losses
    mag_factor = factor(mag, levels = 0:5, ordered = TRUE)      # 9. Mag factor ordered
  ) %>%
  
  # 5. AGGREGATED VALUES
  group_by(yr, st) %>%
  mutate(
    state_annual_fat_total = sum(fat, na.rm = TRUE),            # 10. Total death by state in a year
    state_annual_avg_mag = mean(as.numeric(mag), na.rm = TRUE)  # 11. Average mag by state in a year
  ) %>%
  ungroup() %>%
  
  # 6. EXTRA
  mutate(
    time_of_day = case_when(                                    # 12. Day-time grouping
      lubridate::hour(datetime_utc) >= 6 & lubridate::hour(datetime_utc) < 12 ~ "Morning",
      lubridate::hour(datetime_utc) >= 12 & lubridate::hour(datetime_utc) < 18 ~ "Afternoon",
      lubridate::hour(datetime_utc) >= 18 & lubridate::hour(datetime_utc) < 24 ~ "Evening",
      TRUE ~ "Night"
    )
  )

#glimpse(tornados_clean)

```

---

# Research Questions

## IS THERE A SIGNIFICANT RELATIONSHIP BETWEEN TWO VARIABLES???
The fisrt relation I thougth about is injuries ~ victims (an obvious one) and guided by curiosity, made a bar chart about the proportion of injuries and victims in the tornados

**BARS CHART -> victims_total~injuries~fatalities**
```{r}
# Group by "mag" for clear analysis by plot and give values about (injuries + fatalities)
tornados_resumen <- tornados_clean %>%
  group_by(mag) %>%
  summarise(
    Injuries = sum(inj, na.rm = TRUE),
    Fatalities = sum(fat, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(Injuries, Fatalities), 
    names_to = "Tipo_Victima", 
    values_to = "Cantidad"
  )
tornados_resumen

# 3. Crete chart
ggplot(tornados_resumen, aes(x = factor(mag), y = Cantidad, fill = Tipo_Victima)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Tornado victims by magnitude",
    subtitle = "Total Injuries and Fatalities",
    x = "Tornado magnitude (mag)",
    y = "Total victims",
    fill = "Victim type"
  ) +
  scale_fill_manual(values = c("Fatalities" = "#d95f02", "Injuries" = "#7570b3")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11)
  )
```


After this, to search for relations between variables we went to the correlation matrix of the "eda".

# 1. Analysis of relation ---> "loss_log10" ~ "mag".
**REGRESSION MODEL** ---> Precits losses ("loss_log10") by only knowing the magnitude ("mag")


```{r}

# 1. RELATION CHART
# With factor mag, we make "mag" categorical, not continuious
ggplot(tornados_clean, aes(x = factor(mag), y = loss_log10, fill = factor(mag))) +
  # Add the violin chart
  geom_violin(alpha = 0.5, trim = FALSE) +    # aplha for transparency, trim to cut the violin queues
  # Add a thin black boxplot to see the medians
  geom_boxplot(width = 0.1, color = "black", fill = "white", outlier.shape = NA) +  # we dont represent the outliers
  labs(
    title = "Economic LOSSES(log10) by tornado MAGNITUDE",
    x = "Tornado magnitude",
    y = "Losses (Log10 USD)"
  ) +
  
  scale_fill_brewer(palette = "Reds") + 
  
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11)
  )

```
We can see some relation but we will confirm it with data, not only plot




Definition, Data and Representation of the REGRESSION MODEL

```{r}
# MODEL
modelo_perdidas <- lm(loss_log10 ~ mag, data = tornados_clean)

# MODEL DATA
summary(modelo_perdidas)

# MODEL CHART
ggplot(tornados_clean, aes(x = mag, y = loss_log10)) + # not "factor(mag)" because is linear regression
  geom_jitter(alpha = 0.15, color = "firebrick", width = 0.2) + # adds a little horizontal noise to avoid all the points be the same position. To be more visible
  geom_smooth(method = "lm", color = "black", se = TRUE, size = 1) +  # draws the regression line over the points
  theme_minimal() +
  labs(
    title = "Linear Regression Model: Magnitude vs. Economic Losses",
    subtitle = "Estimated trend line for financial impact (Log10 Scale)",
    x = "Tornado Magnitude (mag)",
    y = "Economic Losses (Log10 USD)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )
```
1Q -> The 25% of residuals are <= than 0.0123
Median -> The 50% of residuals are <= than 0.0123
3Q -> The 75% of residuals are <= than 0.1936



Estimate -> Coefficent value. 
            When "mag=0", "economic losses" = 3.987658.
            And each unit magnitude increases, "economic losses" increase 0.663884
            
Std.Error -> Uncertainty of the "estimate" value.

t value -> "Estimate / Std.Value"

Pr(>|t|) -> Represents the chances of that coefficent being random. If it is < 0.05 is good.
            "***" is the best it can be



RSE -> Squares the errors before dividing to calculate the average error. Punishes the outliers

Multiple R-squared -> The "magnitude" variable explains the 37% of the variability of the "economic loss".                     To can predict the exact value of "economic loss" we would need other variables (63%                      info resting).

                    R squared = 1 - (error of our model / error of only predicting by the mean value)
                    
Adjusted R-squared -> Punishes the number of variables used on the model. If we use variables and they do                     not make part on the predict, this value would increase

F-statistic -> It answers to --- "It is better my model than other model without any parameter?" (in this                  case wothout using mag)
              It measures how better is my model compared to a null one

P - value -> Answers to "Could be my results by chance?" If it is <0.05 is not by chance
            Confirms that mag is not random. It has relation with losses










We can confirm they are related observing the summary and the plot



We check the QUALITY of the model checking some data as the P-VALUE and RSE of summary, calculating MAE (average error) and creating a chart that represents a "termomether" of the error


```{r}
# **P-value** -> ***. Is not luck that relation and "estimate" value

predicciones1 <- predict(modelo_perdidas)

# Calculate MAE (Mean Absolute Error)
mae1 <- mean(abs(tornados_clean$loss_log10 - predicciones1))
# We show the RSE (Residual Standard Error) calculated in the last "summary"
summary1 <- summary(modelo_perdidas)

rse_valor1 <- summary1$sigma
print(paste("Medium Absolute Error -> ", mae1))
print(paste("Residual Standard Error -> ", rse_valor1))


plot(modelo_perdidas, which = 1, col = rgb(0.5, 0, 0.5, 0.1), pch = 16)
  # which = 1 --> selects the chart of "Residuals vs Fited". It answers with 4 differents charts, not important now

  # pch = 16 --> shape of results. In this case, solid circles
```
Regulars - Fitted -> X axis are the predicted values. The Y axis, the errors



Being the RSE > MAE (usually happens), it shows that not all errors are the same. They are irregular, as normal.
RSE is bigger than MAE because there are some cases where it takes a considerably bigger error than common ones (in "outliers" cases), and RSE punish this.

In the error THERMOMETER CHART, we can see some outliers examples that make the RSE bigger



# Analysis of relation ---> "loss_log10" ~ "victims_totals".
**REGRESSION MODEL** ---> Precits losses ("loss_log10") by only knowing the total victims ("victims_total")

Here, we will start to doubt on the correlation matrix data.
In this case, the X axis is not categorical as the "magnitude" was, and started to have some "problems"
So we had some geometric space problems to represent the plot.
Thats why we used "logs" (explanation)
Comparation of data plots with and without log scale

```{r}
sum(tornados_clean %>% 
  filter(victims_total < 0))

```


```{r}
# RELATION CHART

p1 <- ggplot(tornados_clean, aes(x = victims_total, y = loss)) +
  geom_point(alpha = 0.4, color = "#2c3e50") +
  # regression line
  geom_smooth(method = "lm", color = "#e74c3c", se = TRUE) +
  
  labs(
    title = "Relation between Total Victims and Economic Losses",
    subtitle = "Regression Line(lm) about losses",
    x = "Total victims (Injured + Fatalities)",
    y = "Economic Losses"
  ) +
  theme_minimal()


# We make a scatter-plot with
# With loss_log_10 we solve the Y axis problem but not the X
p2<- ggplot(tornados_clean, aes(x = victims_total, y = loss_log10)) +
  geom_point(alpha = 0.4, color = "#2c3e50") +
  
  # regression line
  geom_smooth(method = "lm", color = "#e74c3c", se = TRUE) +
  
  labs(
    title = "Relation between Total Victims and Economic Losses",
    subtitle = "Regression Line(lm) about losses",
    x = "Total victims (Injured + Fatalities)",
    y = "Economic Losses (log10)"
  ) +
  theme_minimal()

# We make a logarithmic variable to see how data changes in a plot representation
# With this new logarithmic variable we solve the X axis
tornados_clean$victims_total_log10 <- log10(tornados_clean$victims_total + 0.001)

p3 <- ggplot(tornados_clean, aes(x = victims_total_log10, y = loss_log10)) +
  geom_point(alpha = 0.4, color = "#2c3e50") +
  
  # regression line
  geom_smooth(method = "lm", color = "#e74c3c", se = TRUE) +
  
  labs(
    title = "Relación entre Víctimas Totales y Pérdidas Económicas",
    subtitle = "Línea de regresión lineal (lm) sobre pérdidas en escala logarítmica",
    x = "Número Total de Víctimas (Heridos + Fallecidos) (Log 10)",
    y = "Pérdidas Económicas (Log10 USD)"
  ) +
  theme_minimal()


modelo_victims <- lm(loss_log10 ~ victims_total_log10, data = tornados_clean)

p1
p2
p3

# Mostramos el modelo 
modelo_perdidas_victimas <- lm(loss_log10 ~ victims_total_log10, data = tornados_clean)
summary(modelo_perdidas_victimas)
```
EXPLANATION about how making a log scale can improve a chart:
  Because of the "outliers", it pushes down all the others "common" cases to the bottom of the chart as we can see in the first case
  
  The log effect, makes the outliers go down and the common ones (smaller) rise up (instead of a tornado be 1000 times worse in losses, with log scale will be 3 units worse)

The last one looks like there is negative victims but actually are the cases with 0 victims. It is because when we make the new "victims_total_log10" we are adding 0.001 tom avoid the log(0) and log(0.001)=-3

ABOUT CHARTS:

We notice that the plot and data of the relation of loss_log10 and victims is high, when in the correlation matrix (loss_log10 ~ vitims_total = 0.2) and (loss ~ victims_total = 0.53), was extrange and ¿why so different?.
We check the "real" correlation using the correlation method "spearman"

```{r}
# "loss" - "victims_total" real correlation
cor(tornados_clean$victims_total, tornados_clean$loss, method = "spearman")
```
¿Why SPEARMAN and not PEARSON?
The PEARSON correlation method is good for perfect data without irregularities and no outliers
The SPEARMAN one, is better for these cases. It doesnt focus on valus but in ranking. This softs the outliers huge values by only asigning them an order value.
It checks if the order of one of the variables is the same (or aprox) of the other variable order.
When closer, better

```{r}
# WHY SPEARMAN

# --- Chart 9: Correlation Matrix of Numerical Variables ---

# Necessary for the correlation matrix
library(reshape2) 

numeric_vars <- tornados_clean %>%
  dplyr::select(    # we select the variables we want in the correlation matrix
    mag, len, wid, inj, fat, 
    loss, victims_total, area_sq_mi, 
    loss_log10
  ) %>%
  drop_na() 

# Calculated matrix with correlation values
cor_matrix <- cor(numeric_vars, method = "spearman")

# Mandatory step to can plot the correlation matrix(it transform the previus table values into 3 varibales (VAR1, VAR2, VALUES)). Easier to use then with ggplot...
melted_cor <- melt(cor_matrix)

# Create the chart
p9 <- ggplot(data = melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +   # adds a white line between values
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +   # round to 2 decimals
  scale_fill_gradient2(
    low = "#0571b0", high = "#ca0020", mid = "white", 
    midpoint = 0, limit = c(-1, 1), space = "Lab", 
    name = "Pearson\nCorrelation"
  ) +
  labs(
    title = "Correlation Matrix of Tornado Metrics",
    subtitle = "Relationship between physical characteristics and impacts",
    x = "", y = "",
    caption = "Source: NOAA | Analysis by Group 15"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), # rotate texts on X axis to not overlap
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.major = element_blank()
  )

print(p9)

# Save it in the Outputs folder
ggsave(
  filename = here("Outputs", "SPEARMAN_tornado_correlation_matrix.png"),
  plot = p9,
  width = 8,
  height = 8,
  units = "in",
  dpi = 1000  
)

```

## Analyzing some strong and interesting relations of the correlation matrix:
# magnitude - loss (0.62)
# magnitude - area_sq_mi (0.49)
# magnitude - victims_total (0.41)
# loss - area_sq_mi(0.5)
# area_sq_mi - len and wide (0.85, 0.93)
# magnitude - boolean_fatalities

#1. RELATION CHART
#2. MODEL, MODEL DATA, MODEL CHART
#3. QUIALITY MODEL CHECK

```{r}
# Now we can see they have considerable and stable values (in the before one "loss" and "loss_log10" were not the same), we confirm our relation thoughts.
```

# 1- Analysis of relation ---> magnitude - loss -> done before

# 2- Analysis of relation ---> magnitude - area_sq_mi
**REGRESSION MODEL** ---> Precits losses ("area_sq_mi") by only knowing the magnitude ("mag")

```{r}

#1.RELATION CHART
# We create a logarithmic value about the destroyed area to avoid the geometric problem of representing in into a chart
tornados_clean$area_sq_mi_log10 <- log10(tornados_clean$area_sq_mi + 0.001)

# We show in the chart the ones that have had > 0 victims with "subset"
ggplot(subset(tornados_clean, area_sq_mi > 0), aes(x = as.factor(mag), y = area_sq_mi_log10, fill = as.factor(mag))) +
  geom_violin(alpha = 0.4, color = "gray40") +  
  geom_boxplot(width = 0.2, color = "black", alpha = 0.7, outlier.shape = NA) + 
  scale_fill_brewer(palette = "Purples") + 
  theme_minimal() + 
  labs(
    title = "Destruction Area by Tornado Magnitude",
    subtitle = "Distribution of affected area (Square Miles in Log10 Scale)",
    x = "Tornado Magnitude (mag)",
    y = "Destroyed Area (Square Miles -> Log10 Scale)",
    fill = "Magnitude"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none"
  )
```
Following this plot, we can see some obvious relation (a clear coefficent of, more mag, more destroyed area)

```{r}
#2. MODEL, MODEL DATA, MODEL CHART

tornados_area_destroyed = tornados_clean%>%filter(area_sq_mi>0) 
  # Eliminated 500 (aprox) tornados of 0 area destroyed.

# MODEL
modelo_destroyed_area <- lm(area_sq_mi_log10 ~ mag, data = tornados_clean)

# MODEL DATA
summary(modelo_destroyed_area)


# MODEL CHART
ggplot(tornados_clean, aes(x = mag, y = area_sq_mi_log10)) +
  geom_jitter(alpha = 0.1, color = "purple4", width = 0.2) + 
  geom_smooth(method = "lm", color = "darkred", se = TRUE, linewidth = 1.2) +
  theme_minimal() +
  labs(
    title = "Linear Regression Model: Magnitude vs. Destroyed Area",
    subtitle = "Estimated trend line using Ordinary Least Squares",
    x = "Tornado Magnitude (mag)",
    y = "Destroyed Area (Log10 Square Miles + 0.001)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )

```
We check the QUALITY of the relation with not only the chart, but also with data

```{r}
#3. QUALITY MODEL CHECK
predicciones2 <- predict(modelo_destroyed_area)

# Calculate MAE ("Mean Absolute Error")
mae2 <- mean(abs(tornados_area_destroyed$area_sq_mi_log10 - predicciones2))

# Show the RSE of the summary done before
summary2 <- summary(modelo_destroyed_area)
rse_valor2 <- summary2$sigma

summary2
print(paste("Mean Absolute Error -> ", mae2))
print(paste("Residual Standard Error -> ", rse_valor2))

# Residuals chart
plot(modelo_destroyed_area, which = 1, col = rgb(0.5, 0, 0.5, 0.1), pch = 16)

```

In this case it also shows they are related by plot and summary data.
Is true that here the MAE and RSE iss bigger, but we have to take note that it is only one independent variable. It "contains" the 29% info. we need to get the perfect prediction.
It is related. If we had more variables on the indepentendt side, we would tight the error.

The values that confirms this relation about more mag ---> more area_destroyed is not by chance are there, like p-value or t-value.

Also, we did a case where we only had tornados with >0 area destroyed, but actually, these cases are "normal" tornado cases (there are 500 of 68693) that we should not exclude in a linear regression.
We could do a logisitc regression to predict if they are going to be <0 or >0 but is only a few cases of the total so we wont do it (we could have adapted it to setting a threshold, and if >threshold is high destroyed area and <threshold low, but we will do that with tornado magnitude)


# Now we will do a deeper analysis in the "victims_total" variable with the others:

# 3- Analysis of relation ---> victims_total ~ magnitude 
**REGRESSION MODEL** ---> Predicts total victims("total_victims_log10") by only knowing the magnitude ("mag")

# 4- Analysis of relation ---> victims_total (0.28) ~ area_sq_mi 
**REGRESSION MODEL** ---> Predicts total victims ("total_victims_log10") by only knowing the area destroyed ("area_sq_mi_log10")

# 4- Analysis of relation ---> victims_total (0.28) ~ loss_log10
**REGRESSION MODEL** ---> Predicts total victims ("total_victims_log10") by only knowing the losses ("loss_log10")

#### We filter the dataset to ONLY take the tornados WITH VICTIMS (are not that common in the dataset)
```{r}
# The variable of "victims_totals" is more delicate. There are many cases with 0 victims or 0 fatalities
victims_data <- tornados_clean %>%
  mutate(categoria = case_when(
    is_deadly ~ "Deadly",
    victims_total > 0 ~ "Injured and deadly",
    TRUE ~ "No victims"
  )) %>%
  count(categoria)

victims_data

ggplot(victims_data, aes(x = categoria, y = n, fill = n)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = n),  # Shows n value
    vjust = -0.5,    # Adjust text to can be seen clearly
    size = 3.5,      # Text size
    color = "black"  # Text color
  ) +
  scale_fill_gradient(
    low = "lightblue",   # Low values color
    high = "darkblue"    # High values color
  ) +
  labs(
    title = "Tornados by category of victim type",
    x = "",
    y = "Tornados count"
  ) +
  theme_minimal()
```
As we can see ONLY 7964 of 68693 tornados have some victims.

Now, we will make 3 models with each vairbale individually of these tornado victim cases, and then make a SECOND model using these 3 independent variables to see the difference.

We will do it separately to observe the difference
```{r}
# We eliminate those tornados without area destroyed
victims_only <- tornados_clean %>% filter(victims_total >=1)
# >= 1 victims
count(victims_only)

# We make the charts of the relation X and Y axis of the analysed variables

p1 <- ggplot(victims_only, aes(x = factor(mag), y = victims_total_log10, fill = factor(mag))) +
  geom_violin(alpha = 0.5, trim = FALSE, color = "grey") +
  geom_boxplot(width = 0.2, color = "black", alpha = 0.7, outlier.alpha = 0.1) +
  scale_fill_brewer(palette = "Purples") +
  labs(title = "Victims by tornado magnitude (mag)", x = "magnitude", y = "Total Victims(log10)") +
  theme(legend.position = "none")




# Showed in the next cell code all together
p2 <- ggplot(victims_only, aes(x = area_sq_mi_log10, y = victims_total_log10)) +
  geom_point(alpha = 0.3, color = "orange2") +
  geom_smooth(method = "lm", color = "darkred") +
  labs(title = "Victims b destroyed area (area_sq_mi_log10)", x = "Destroyed area (log10)", y = "Total Victims(log10)")

p3 <- ggplot(victims_only, aes(x = loss_log10, y = victims_total_log10)) +
  geom_jitter(alpha = 0.3, color = "black", width = 0.2) + 
  geom_smooth(method = "lm", color = "darkred", se = TRUE, linewidth = 1.2) + # Nota: 'size' se actualizó a 'linewidth' en versiones nuevas de ggplot
  theme_minimal() +
  labs(
    title = "Linear Regression Model: Losses vs. Victims Total",
    subtitle = "Estimated trend line using Ordinary Least Squares",
    x = "Losses (log10)",
    y = "Victims Total (Log10 + 0.001)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )

p1

```


```{r}
# 2. MODEL, MODEL DATA, MODEL CHART

# MODELS
modelo_magnitude_victim <- lm(victims_total_log10 ~ mag, data = victims_only)
modelo_area_sq_mi_log10_victim <- lm(victims_total_log10 ~ area_sq_mi_log10, data = victims_only)
modelo_losses_victim <- lm(victims_total_log10 ~ loss_log10, data = victims_only)

# MODEL DATA
summary3 <- summary(modelo_magnitude_victim)
summary4 <- summary(modelo_area_sq_mi_log10_victim)
summary5 <- summary(modelo_losses_victim)


# MODEL CHART
# MAGNITUDE-VICTIM
# Without the "factor(mag)", to a linear regression
p4 <- ggplot(victims_only, aes(x = mag, y = victims_total_log10)) +
  geom_jitter(alpha = 0.3, color = "purple4", width = 0.2) + 
  geom_smooth(method = "lm", color = "darkred", se = TRUE, linewidth = 1.2) + # Nota: 'size' se actualizó a 'linewidth' en versiones nuevas de ggplot
  theme_minimal() +
  labs(
    title = "Linear Regression Model: Magnitude vs. Victims Total",
    x = "Tornado Magnitude (mag)",
    y = "Victims Total (Log10 + 0.001)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )





print("Summary of model VICTIMS-MAGNITUDE ")
summary3
print("Summary of model VICTIMS-AREA_SQ_MI_LOG10 ")
summary4
print("Summary of model VICTIMS-LOSSES_LOG_10 ")
summary5

p4
p2
p3

# All the pontis of Axis Y == 0 is because they have only 1 victim (log(1)=0)
count(victims_only%>%filter(victims_total==1))

```
EXPLICATION:
All of them 3 have values that make us think they are related and are significatively important.



```{r}
plot(modelo_magnitude_victim, which = 1, col = rgb(0.5, 0, 0.5, 0.1), pch = 16)
plot(modelo_area_sq_mi_log10_victim, which = 1, col = rgb(0.5, 0, 0.5, 0.1), pch = 16)
plot(modelo_losses_victim, which = 1, col = rgb(0.5, 0, 0.5, 0.1), pch = 16)

```
EXPLICATION ERROR THERMOMETER:
These last 3 charts shows us if the model is "optimist" or not, by showing us how it fails. If it predict is higer or lower value than the real one



# 5- Analysis of relation ---> victims_total - magnitude + area_sq_mi + log_loss10
**MULTI-REGRESSION MODEL** ---> Precits total victims("total_victims_log10") by only knowing the magnitude ("mag")

We have data to analyze the model individually to can compare next when we make a multi-variable model
```{r}
# MODEL
modelo_multivariable <- lm(victims_total_log10 ~ mag + area_sq_mi_log10 + loss_log10, data = victims_only)
summary6<-summary(modelo_multivariable)
summary6

print("Error summary3, VICTIMS-MAGNITUDE")
summary3$sigma

print("Error summary4 VICTIMS-AREA_SQ_MI_LOG10")
summary4$sigma

print("Error summary5 VICTIMS-LOSSES_LOG_10")
summary5$sigma

print("Error multivariable")
summary6$sigma

# MODEL CHART
victims_only$predicted <- predict(modelo_multivariable)

ggplot(victims_only, aes(x = predicted, y = victims_total_log10)) +
  geom_point(alpha = 0.2, color = "purple4") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  theme_minimal() +
  labs(title = "Valores observados vs. predichos",
       subtitle = "Modelo: victims_total_log10 ~ mag + area_sq_mi_log10 + loss_log10",
       x = "Valor predicho por el modelo",
       y = "Valor real (victims_total_log10)")


plot(modelo_multivariable, which = 1, col = rgb(0.5, 0, 0.5, 0.1), pch = 16)


coefs <- tidy(modelo_multivariable, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

ggplot(coefs, aes(x = term, y = estimate)) +
  geom_point(size = 3, color = "darkred") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  theme_minimal() +
  labs(title = "Coeficients of the multivariable model",
       x = "", y = "Estimmation (confidence interval of 95%)")

```
EXPLICATION:

What we can see is that the multivariable model is better than the 3 by their own.

MULTIPLE R SQUARED -> It explains the 38% of the variability of "citims_total_log10" when the others explains only the 29%(mag) 17%(area_sq_mi) 27%(loss_log10)

"The reason the final R² is 38% instead of the sum of 29% + 17% + 27% is due to multicollinearity. This means that our three predictor variables are not fully independent from each other — they are physically related phenomena. A tornado with higher magnitude tends to also destroy a larger area, and a larger destroyed area tends to produce higher economic losses. Because of this, when we combine the three variables in one model, they partially explain the same portion of the variance in victims, so their contributions overlap rather than stack up. Each variable only gets credit for what it explains exclusively, beyond what the other two already captured."



ADJUSTED R-SQUARED -> If any of the added variables to the model wont be useful or important, it woul have punish it and decrease, but all are significative and improve the model

RSE -> Has the lowest RSE of all the models

        [1] "Error summary3, VICTIMS-MAGNITUDE"
        [1] 0.4703687   --->   10^0.4704  # Error del modelo mag → ~2.95 víctimas
        [1] "Error summary4 VICTIMS-AREA_SQ_MI_LOG10"
        [1] 0.5110098   --->   10^0.5110  # Error del modelo area → ~3.24 víctimas
        [1] "Error summary5 VICTIMS-LOSSES_LOG_10"
        [1] 0.4774674   --->   10^0.4775  # Error del modelo loss → ~3.00 víctimas

        
        [1] "Error multivariable"
        [1] 0.4411276   --->   10^0.4411  # Error multivariable → ~2.76 víctimas


The coefficent values (last chart explanation and summary)

Overall, the multivariable model achieves a better balance between explanatory power and prediction error. It explains 9 percentage points more variance than the best individual model (mag at 29%) while reducing the average prediction error from ~2.95 victims to ~2.76 victims in real scale. However, the moderate R² of 38% suggests that other factors not present in the dataset, would be needed to further improve predictions.




Now with "anova" we ask ourselves:
"Is the multivariable model significantly better than the simple model, or could the improvement be due to pure chance?"
```{r}
# Other fast way to compare models
anova(modelo_magnitude_victim, modelo_multivariable)
anova(modelo_area_sq_mi_log10_victim, modelo_multivariable)
anova(modelo_losses_victim, modelo_multivariable)
```
EXPLICATION of results:

RSS (Residual Sum of Squares) → Total sum of all squared errors. The lower, the better the model fits the data.

Sum of Sq → The difference in RSS between the two models. How much error was reduced by adding the extra variables.

F → Same as the F-statistic seen before, measures whether that error reduction is relevant or trivial.
    F = (Sum of Sq / Df) / (RSS_complejo / Res.Df)
          Sum of Sq → Error reduction when adding variables
          Df → Number of added variables
          RSS_complex → Residual error of the more complex model
          Res.Df → Degrees of freedom of the complex model

              F < 4 → little improve, probably not significative
              F > 4 → starts to be significative (with big datasets)
              F > 10 → clearly significative
              F > 100 → big improvement
              F > 1000 → huge improvement

Pr(>F) → The p-value. If it is lower than 0.05 the improvement is statistically significant.

```{r}
# We compare 3 individuals models as before with the multi-variable one
library(broom)
library(dplyr)

comparacion <- bind_rows(
  glance(modelo_magnitude_victim)  %>% mutate(modelo = "Magnitude"),
  glance(modelo_area_sq_mi_log10_victim) %>% mutate(modelo = "Area_sq_mi_log10"),
  glance(modelo_losses_victim)     %>% mutate(modelo = "loss_log10"),
  glance(modelo_multivariable)     %>% mutate(modelo = "Multivariable")
) %>%
  select(modelo, r.squared, adj.r.squared, AIC, BIC)

comparacion

ggplot(comparacion, aes(x = modelo, y = adj.r.squared, fill = modelo)) +
  geom_col() +
  geom_text(aes(label = round(adj.r.squared, 3)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Comparation of 'adjusted square-R ajustado' between models",
       x = "", y = "R-cuadrado ajustado") +
  theme(legend.position = "none")
```
EXPLANATION:

Finally we can check the AIC and BIC (these ones are when LOWER -> BETTER).
These data exist to compare models fairly: rewards good fit while penalizing unnecessary complexity.

AIC (Akaike Information Criterion) → Penalizes each added variable in a moderate way. It favors models that balance good fit with simplicity.

BIC (Bayesian Information Criterion) → Penalizes more strongly than AIC, especially when you have a large amount of data. It is more strict when it comes to accepting new variables.


AIC = (how poorly the model fits) + 2 × (number of variables)
BIC = (how poorly the model fits) + ln(n) × (number of variables)
Where n is the number of observations (rows in your dataset).

Since we have thousands of tornadoes, ln(n) is much larger than 2, which is why BIC penalizes more strongly than AIC in your case. The more data you have, the stricter BIC becomes when accepting new variables.



The multivariable model has the lowest AIC and BIC of all, confirming that adding the three variables together is worth it even after penalizing for added complexity.

```{r}
library(rsample)
# Now we are going to do a logistic regression.
# Having the data we have
tornados_clean <- tornados_clean %>%
  mutate(mag_alta = ifelse(mag >= 3, 1, 0))

# Dataset de entrenamiento y de test
set.seed(123) 
split_objeto   <- initial_split(tornados_clean, prop = 0.75, strata = mag_alta)
tornados_train <- training(split_objeto)
tornados_test  <- testing(split_objeto)



modelo_logistico <- glm(mag_alta ~ area_sq_mi_log10 + loss_log10 + victims_total_log10,
                         data = tornados_train, family = binomial)
summary(modelo_logistico)
```
EXPLANATION:

These results are so similar to the summaries of the linera regression models. But this is a "LOGISTC REGRESSION" and we interpret them in a different way.

In this case we dont treat the coefficents as Y rising continious values, we treat them as probabilities (between 0 and 1).
For that we work with odds instead of probabilities:
      odds = probability / (1 - probability)
      
      Example: if the probability of a tornado being high magnitude is 0.8 (80%):
      odds = 0.8 / (1 - 0.8) = 0.8 / 0.2 = 4
      This means "4 times more likely to happen than not to happen."
      
To pass these odds to probabilities:
      exp(0.94907)  # area_sq_mi_log10
      exp(0.80720)  # loss_log10
      exp(0.43720)  # victims_total_log10
      
For example exp(0.949) ≈ 2.58, meaning that for every unit increase in area_sq_mi_log10, the probabilities of it being high magnitude are multiplied by 2.58.


Null deviance (19272) -> the error of the simplest possible model, with no predictor variables, only the intercept


Residual deviance (10956) -> the error of your model with the 3 variables



```{r}
# Instala el paquete si no lo tienes: install.packages("sjPlot")
library(sjPlot)

# Este comando dibuja los Odds Ratios de forma automática
plot_model(modelo_logistico, vline.color = "red", sort.est = TRUE)
```
Visual representation about the lasts results.





```{r}
# 2. Calculate the individual predictions

# ggopredict -> 
# For each variable, it fixes the other two at their mean value and varies only that variable to see how the predicted probability of mag_alta changes. It's like saying "holding everything else constant, what happens if I only change the area?"


pred_area    <- ggpredict(modelo_logistico, terms = "area_sq_mi_log10")
pred_loss    <- ggpredict(modelo_logistico, terms = "loss_log10")
pred_victims <- ggpredict(modelo_logistico, terms = "victims_total_log10")

# 3. Create the chart for area_sq_mi_log10
g1 <- ggplot(pred_area, aes(x = x, y = predicted)) +
  geom_line(color = "#1f77b4", size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#1f77b4", alpha = 0.15) +
  labs(title = "Efecto del Área", x = "Área (log10 sq mi)", y = "Probabilidad Mag Alta") +
  theme_minimal(base_size = 12) +
  ylim(0, 1) # Importante: fija el eje Y de 0 a 1 para ver la probabilidad real

# 4. Creamos el gráfico para las PÉRDIDAS (Loss)
g2 <- ggplot(pred_loss, aes(x = x, y = predicted)) +
  geom_line(color = "#e377c2", size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#e377c2", alpha = 0.15) +
  labs(title = "Efecto de las Pérdidas", x = "Pérdidas (log10 $)", y = "") +
  theme_minimal(base_size = 12) +
  ylim(0, 1)

# 5. Creamos el gráfico para las VÍCTIMAS
g3 <- ggplot(pred_victims, aes(x = x, y = predicted)) +
  geom_line(color = "#2ca02c", size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#2ca02c", alpha = 0.15) +
  labs(title = "Efecto de las Víctimas", x = "Total Víctimas (log10)", y = "Probabilidad Mag Alta") +
  theme_minimal(base_size = 12) +
  ylim(0, 1)

# 6. Los juntamos de forma espectacular gracias a patchwork
(g1 + g2) / g3 + plot_annotation(title = "Marginal effects of the Logisitic Mode",
                                 subtitle = "Predicted probability of a high mag tornado (>= 3)")
```
EXPLANATION:
We can confirm what we saw before.
The "area_sq_mi_log10" and "loss_log10" are the most important variables for the predicition of a tornado magnitude. As we can see, they both have a great change on the "probability" of being "high mag" when their values reach certain point.
The "victims_total_log10" is softer and does not change as the other two





```{r}

# Guardamos las predicciones en el dataframe de TEST para no mezclar datos
tornados_test$pred_prob <- predict(modelo_logistico, newdata = tornados_test, type = "response") # response is to the model return probabilities between 0 and 1 and not the raw logit
tornados_test$pred_clase <- ifelse(tornados_test$pred_prob > 0.5, 1, 0)

# 3. Graficar la matriz de confusión
eval <- evaluate(
  data = tornados_test, 
  target_col = "mag_alta", 
  prediction_cols = "pred_clase", # En plural
  type = "binomial"                # ¡Este es el que faltaba!
)

plot_confusion_matrix(eval)
```

```{r}
# Now we will analyze the ROC curve. This curve represents what happens to the model when we change the "threshold".
# Y axis --> Recall
# X axis --> Fake Positives

# It has different methods to calculate it.
# In this case where we want to priorize the Recall (we dont want FAKE NEGATIVES) we use "YOUDEN" that is "Recall + Specifity" (we maximize the TRUE POSITIVES and minimize the FAKE NEGATIVES)

# Moving the threshold always will improve one thing and worse other.

library(caret)

reporte_automatico <- confusionMatrix(
  data = as.factor(tornados_test$pred_clase),       # ¡Cambiado a _test!
  reference = as.factor(tornados_test$mag_alta),    # ¡Cambiado a _test!
  positive = "1"
)

# 3. Extraemos e imprimimos solo las métricas clave que necesitas para tu reporte:
print("--- MÉTRICAS AUTOMÁTICAS DEL MODELO LOGÍSTICO ---")
print(paste("(Accuracy):", round(reporte_automatico$overall["Accuracy"] * 100, 2), "%"))
print(paste("Precision):", round(reporte_automatico$byClass["Precision"] * 100, 2), "%"))
print(paste("(Recall / Sensitivity):", round(reporte_automatico$byClass["Sensitivity"] * 100, 2), "%"))
print(paste("F1-Score (Balance):", round(reporte_automatico$byClass["F1"], 4)))

```
EXPLANATION:

Accuracy (96.15%) -> "Of all predictions made, how many were correct overall?"
                    Not always a high accuracy is a good model, as this case
                    
Precision (70.62%) -> "Of everything I predicted as positive, how many actually were positive?"
                    It measures the quality of the "alarms"
                    
Recall (33.33%) -> "Of all the actual positives, how many did I correctly detect?"
                    In this case we have a "LAZY" model (we will explain this)
                    
F1-Score (0.4529) -> "The harmonic balance between Precision and Recall."

```{r}
library(pROC)

# 1. Calculate the ROC curve
roc_objeto <- roc(
  response  = tornados_test$mag_alta,
  predictor = tornados_test$pred_prob
)

# 2. Observe the AUC (Area Under the Curve)
print(paste("AUC:", round(auc(roc_objeto), 4)))

# 3. Encontrar el threshold óptimo (método de Youden)
threshold_optimo <- coords(roc_objeto, "best", best.method = "youden")
print(threshold_optimo)

# 4. Graficar la curva ROC
plot(roc_objeto, 
     col  = "blue", 
     main = "ROC Curve - Logistic Model Tornados",
     print.auc = TRUE)

# 5. Marcar el punto óptimo en la curva
points(threshold_optimo$specificity, 
       threshold_optimo$sensitivity, 
       col = "red", pch = 19, cex = 1.5)

```
EXPLANATION:

Recall (Y axis) → of the actual positives, how many does it detect
Specificity (X axis) → of the actual negatives, how many does it detect

ROC Curve -> The ROC curve goes through all possible thresholds (from 0 to 1), and for each one it calculates:



Each threshold generates one blue point, and all those points together form the curve.

The red point is the zone where you're getting the maximum possible Recall in exchange for the smallest sacrifice in Specificity

AUC (Area Under the Curve) -> 
                        Measures how well a model can distinguish between the two classes, across all                             possible thresholds. It represents the probability that the model assigns a higher                         predicted probability to a randomly chosen positive case than to a randomly chosen                         negative case


WHAT WE ARE GOING TO TRY?
In this tornado cases the Recall is much important than other values, as we want specially to detect the DANGEROUS ones. For that, we try to balance the model and the files with a similar number of classes



```{r}
# We balance it
# 1. We separate the "High" de los "Low" in the train set
altos <- tornados_train %>% filter(mag_alta == 1)
bajos <- tornados_train %>% filter(mag_alta == 0)

n_altos <- nrow(altos)
n_bajos <- nrow(bajos)
porcentaje_bajos <- n_bajos / (n_altos+n_bajos)
print(paste("This is the '%' of LWOW MAGNITUDE", round(porcentaje_bajos * 100, 2), "%"))
# 95% de datos de tornados que son MAGNITUD BAJA


# 2. We select same "Lows" as "High" we have
set.seed(123)
bajos_recortados <- bajos %>% sample_n(nrow(altos)) # take as much files of high mag as the files og low mag

# 3. We join them. They are in a 50% 50% proportion
train_balanceado_manual <- bind_rows(altos, bajos_recortados)

# 4. Train the model with this new data
modelo_balanceado_manual <- glm(mag_alta ~ area_sq_mi_log10 + loss_log10 + victims_total_log10,
                     data = train_balanceado_manual, family = binomial)

summary(modelo_balanceado_manual)
```

```{r}
# We follow the same before process
tornados_test$pred_prob <- predict(modelo_balanceado_manual, newdata = tornados_test, type = "response")
tornados_test$pred_clase <- ifelse(tornados_test$pred_prob > 0.5, 1, 0)

# 3. Graficar la matriz de confusión
eval <- evaluate(
  data = tornados_test, 
  target_col = "mag_alta", 
  prediction_cols = "pred_clase", # En plural
  type = "binomial"                # ¡Este es el que faltaba!
)

plot_confusion_matrix(eval)
```


```{r}
reporte_automatico_balanceado <- confusionMatrix(
  data = as.factor(tornados_test$pred_clase),       # ¡Cambiado a _test!
  reference = as.factor(tornados_test$mag_alta),    # ¡Cambiado a _test!
  positive = "1"
)

# 3. Extraemos e imprimimos solo las métricas clave que necesitas para tu reporte:
print("--- MÉTRICAS AUTOMÁTICAS DEL MODELO LOGÍSTICO ---")
print(paste("Exactitud Global (Accuracy):", round(reporte_automatico_balanceado$overall["Accuracy"] * 100, 2), "%"))
print(paste("Precisión (Precision):", round(reporte_automatico_balanceado$byClass["Precision"] * 100, 2), "%"))
print(paste("Sensibilidad (Recall / Sensitivity):", round(reporte_automatico_balanceado$byClass["Sensitivity"] * 100, 2), "%"))
print(paste("F1-Score (Balance):", round(reporte_automatico_balanceado$byClass["F1"], 4)))
```
EXPLANATION OF RESULTS:
In this tornado case, this model would be better because it is more reliable when it predicts a HIGH MAGNITUDE tornado. Is true that it fails more than the other in the other hand: FAKE POSITIVES (it predicts is high but is low)
Is a different model result but better in this special tornado cases


```{r}
# CONCLUSIONS
# At this moment we only thougth about the balance, but we did not thjink about the "threshold" of -->
# > 0.5 = HIGH
# < 0.5 = LOW

# So we started to think about changing that "threshold", and in our tornado case, is worse prediciting a LOW MAGNTIUDE tornado when actually is HIGH MAGNITUDE because we wont activate ways of prevention.

# In the other case, detecting a HIGH MAGNITUDE tornado, when is LOW MAGNITUDE one is not as bad.

# So in conclusion we prefer FAKE POSITIVES than FAKE NEGATIVES. For that, it would be a proper option lower the threshold.

```

```{r}
# We now try the same. Balanced 50% 50% but with the optimous threshold calculated by the ROC curve

# 3. Aplicar el nuevo threshold en vez del 0.5
tornados_test$pred_clase_nueva <- ifelse(
  tornados_test$pred_prob > threshold_optimo$threshold, 1, 0
)

# 4. Comparar métricas antiguas (0.5) vs nuevas (threshold óptimo)
reporte_nuevo <- confusionMatrix(
  data      = as.factor(tornados_test$pred_clase_nueva),
  reference = as.factor(tornados_test$mag_alta),
  positive  = "1"
)

print("--- CON THRESHOLD 0.5 ---")
print(paste("Recall:",    round(reporte_automatico$byClass["Sensitivity"] * 100, 2), "%"))
print(paste("Precision:", round(reporte_automatico$byClass["Precision"]   * 100, 2), "%"))
print(paste("F1:",        round(reporte_automatico$byClass["F1"], 4)))

print("--- CON THRESHOLD ÓPTIMO ---")
print(paste("Recall:",    round(reporte_nuevo$byClass["Sensitivity"] * 100, 2), "%"))
print(paste("Precision:", round(reporte_nuevo$byClass["Precision"]   * 100, 2), "%"))
print(paste("F1:",        round(reporte_nuevo$byClass["F1"], 4)))

```
EXPLANATION: Why F1 with optimum threshold is lower than the 0.5 threshold.

Lowering the threshold increased Recall a lot (from 33% to 84%) but reduced Precision (from 71% to 22%), so the F1-score went down. This happens because F1 punishes a low Precision more than it rewards a high Recall. Even so, in this context it may still be the better choice, since missing a dangerous high-magnitude tornado is worse than triggering a false alarm